Concurrent Programming with CXL

### 2020-06-07 book/paper.tex
Latest

While spinlocks suffer from starvation, it is a uniform random
process and each process has an equal chance of entering the critical
section.  Thus the probability of starvation is exponentially vanishing.
Unfortunately, such is not the case for the
reader/writer solutions that we presented thus far.
Consider this scenario: there are two readers and one writer.  One reader
is in the critical section while the writer is waiting.  Now the
second reader tries to enter and is able to.  The first reader leaves.
We are now in a similar situation as the initial state with one reader
in the critical section and the writer waiting, but it is not the same
reader.  Unfortunately for the writer, this scenario can repeat itself
indefinitely.  So even if neither reader was in the critical section
all of the time, and the second reader arrived well after the writer,
the writer never had a chance.
In this chapter, we will present a version of a reader/writer lock
implementation that solves this type of starvation, but won't eliminate
the type of starvation.

@@

>>>>

While spinlocks suffer from starvation, it is a uniform random
process and each process has an equal chance of entering the critical
section.  Thus the probability of starvation is exponentially vanishing.
Unfortunately, such is not the case for the
reader/writer solutions that we presented thus far.
Consider this scenario: there are two readers and one writer.  One reader
is in the critical section while the writer is waiting.  Now the
second reader tries to enter and is able to.  The first reader leaves.
We are now in a similar situation as the initial state with one reader
in the critical section and the writer waiting, but it is not the same
reader.  Unfortunately for the writer, this scenario can repeat itself
indefinitely.  So even if neither reader was in the critical section
all of the time, and the second reader arrived well after the writer,
the writer never had a chance.

In this chapter, we will present a version of a reader/writer lock
implementation that solves this type of starvation, but won't eliminate
the type of starvation.
In particular, the probability that either a reader or writer gets starved
will be exponentially diminishing.
We shall call such a solution \emph{fair} (although it does not quite
match the usual formal nor vernacular concepts of fairness).

It is difficult if not impossible to make the reader/writer lock solution of
Figure~\ref{fig:rw2lock} fair.  However, the split binary semaphore solution of
Figure~\ref{fig:RWsplitsema} provides much control over whom to pass baton
to.
Figure~\ref{fig:RWfair} contains a version of Figure~\ref{fig:RWsplitsema}
that is a fair solution to the reader/writer problem.  There are two changes
compared to the original version:

\begin{enumerate}
\item When a reader tries to enter the critical section, it yields not only
if there are writers in the critical section, but also if there are writers
waiting to enter the critical section;
\item There are two versions of \texttt{V\_one()}.  In particular, when the
last reader leaves the critical section, it preferentially passes the baton
to a waiting writer rather than to a waiting reader.  Similarly,
when a writer leaves the critical section, it preferentially passes
the baton to a waiting reader rather than to a waiting writer.
\end{enumerate}

The net effect of this is that if there is contention between readers and
writers, than readers and writers end up alternating entering the critical
section.  While readers can still starve other readers and writers can still
starve other writers, they can not starve one another.  And contention for
a semaphore is resolved probabilistically, each reader and each writer will
eventually have a chance to enter the critical section almost surely.

@@

Tony Hoare, who came up with the concept of split binary semaphores, devised
an abstraction of the concept in a programming language paradigm called
\emph{monitors}.
